A Framework for Substructural Type Systems

### 2021-09-20 tex/cpp21/env.tex
[ cpp21 ] compilation and bibliography fixes

\begin{lemma}
  \cref{def:lr-env} is sound and complete for the following syntax.
  \begin{align*}
    \begin{prooftree}[comb,center=false]
      \hypo{I^*}
      \infer1{\alr{} : {} \env\V {\cdot}}
    \end{prooftree}
    &&
    \begin{prooftree}[comb,center=false]
      \hypo{\rho : {} \env\V \Delta_l}
      \hypo{\sep}
      \hypo{\sigma : {} \env\V \Delta_r}
      \infer3{\alr{\rho,\sigma} : {} \env\V \Delta_l, \Delta_r}
    \end{prooftree}
    &&
    \begin{prooftree}[comb,center=false]
      \hypo{\gr r\cdot\plr{M : {} \sdtstile{}\V A}}
      \infer1{\alr{M} : {} \env\V \gr rA}
    \end{prooftree}
  \end{align*}
\end{lemma}
\begin{proof}
  This replaces \cref{thm:construct-env}.
\end{proof}

@@

The requirements for identity and composition of environments look a bit like
the unit and lift of a Kleisli triple.
\todo{Change from equality to inequality}

\begin{lemma}[Identity environment]\label{thm:env-id}
  Given a function $\mathrm{vr} : \forallb{{\sqni} \dotto \V}$, for any
  $\Gamma$ we have an environment of type $\Gamma \env\V \Gamma$.
\end{lemma}
\begin{proof}
  Let $\gr\Psi$ be the identity map, which clearly satisfies
  $\grP = \grP\gr\Psi$.
  When acting on a variable, the equation $\grPprime = \grQprime\gr\Psi$ means
  that $\grPprime = \grQprime$, so we want, from a variable of type
  $\grPprime\gamma \sqni A$, a value of type $\V\,\grPprime\gamma\,A$, which
  we can get from $\mathrm{vr}$.
\end{proof}

\begin{lemma}\label{thm:env-comp-lemma}
  Given an environment $\rho : \Gamma \env\U \Delta$ for which we have, for any
  $\grPprime$ and $\grQprime$ such that
  $\grPprime = \grQprime\plr{\rho.\gr\Psi}$, we have a function
  $\mathrm{lift}_\rho :
  \forallb{\V\,\grQprime\delta \dotto \W\,\grPprime\gamma}$,
  we can map environments of type $\Delta \env\V \Theta$ into environments of
  type $\Gamma \env\W \Theta$.
\end{lemma}
\begin{proof}
  Let $\rho$ be as in the statement, and let $\sigma : \Delta \env\V \Theta$.
  For the environment we are constructing, let
  $\gr\Psi \coloneqq \sigma.\gr\Psi; \rho.\gr\Psi$, noting that
  $\grP = \grQ\plr{\rho.\gr\Psi} =
  \plr{\grR\plr{\sigma.\gr\Psi}}\plr{\rho.\gr\Psi}$.
  For the action on variables, we are given $\grPprime = \grRprime\gr\Psi$ with
  $\grRprime\theta \sqni A$.
  We can immediately apply the action of $\sigma$, giving us a value of type
  $\V\,\plr{\grRprime\plr{\sigma.\gr\Psi}}\,A$.
  We note that
  $\grPprime = \plr{\grRprime\plr{\sigma.\gr\Psi}}\plr{\rho.\gr\Psi}$, and
  apply $\mathrm{lift}_\rho$ to get the desired value.
\end{proof}

\begin{corollary}[Composition of environments]\label{thm:env-comp}
  Given a function
  $\mathrm{lift} : \plr{\rho : \grP\gamma \env\U \grQ\delta} \to
  \forall \grPprime, \grQprime.~\grPprime = \grQprime\plr{\rho.\gr\Psi} \to
  \forallb{\V\,\grQprime\delta \dotto \W\,\grPprime\gamma}$, then we can
  compose environments of types $\Gamma \env\U \Delta$ and
  $\Delta \env\V \Theta$ into an environment of type $\Gamma \env\W \Theta$.
\end{corollary}

\todo{What do these choices mean?}
\begin{example}
  We can derive the following instances of environment composition.
  \begin{itemize}
    \item If $\U = \V = \W = {\sqni}$, then $\mathrm{lift}$ is given by the
      action of the renaming $\rho$ on variables.
      This allows us to derive composition of renamings.
    \item More generally, if $\V = {\sqni}$ and $\U = \W$, we can still use
      the action of the environment $\rho$.
      This means that renamings post-compose with any other sort of environment.
    \item If $\V = \W = {\vdash}$, then $\mathrm{lift}$ is given by a
      syntactic traversal.
      For example, if $\U = {\sqni}$, we need the action of renaming on terms
      to show that a renaming followed by a substitution composes to a
      substitution.
      If $\U = {\vdash}$, then the action of substitution on terms gives us that
      substitutions compose.
    \item More generally, if $\V = {\vdash}$ and we have a semantics from
      $\U$ to $\W$, then $\mathrm{lift}$ can be given by the semantic traversal
      of terms.
  \end{itemize}
\end{example}

>>>>

\begin{lemma}
  \cref{def:lr-env} is sound and complete for the following syntax.
  \begin{displaymath}
    \begin{prooftree}[comb,center=false]
      \hypo{I^*}
      \infer1{\alr{} : {} \env\V {\cdot}}
    \end{prooftree}
    \qquad
    \begin{prooftree}[comb,center=false]
      \hypo{\rho : {} \env\V \Delta_l}
      \hypo{\sep}
      \hypo{\sigma : {} \env\V \Delta_r}
      \infer3{\alr{\rho,\sigma} : {} \env\V \Delta_l, \Delta_r}
    \end{prooftree}
    \qquad
    \begin{prooftree}[comb,center=false]
      \hypo{\gr r\cdot\plr{M : {} \sdtstile{}\V A}}
      \infer1{\alr{M} : {} \env\V \gr rA}
    \end{prooftree}
  \end{displaymath}
\end{lemma}
\begin{proof}
  This replaces \cref{thm:construct-env}.
\end{proof}

@@

The requirements for identity and composition of environments look a bit like
the unit and lift of a Kleisli triple.

\begin{lemma}[Identity environment]\label{thm:env-id}
  Given a function $\mathrm{vr} : \forallb{{\sqni} \dotto \V}$, for any
  $\Gamma$ we have an environment of type $\Gamma \env\V \Gamma$.
\end{lemma}
\begin{proof}
  Let $\gr\Psi$ be the identity map, which clearly satisfies
  $\grP \leq \grP\gr\Psi$.
  When acting on a variable, the inequality $\grPprime \leq \grQprime\gr\Psi$
  means that $\grPprime \leq \grQprime$, so the variable of type
  $\grQprime\gamma \sqni A$ can be coerced to have type
  $\grPprime\gamma \sqni A$.
  From this point, $\mathrm{vr}$ gives us a value of type
  $\V\,\grPprime\gamma\,A$, as required.
\end{proof}

\begin{lemma}\label{thm:env-comp-lemma}
  Given an environment $\rho : \Gamma \env\U \Delta$ for which we have, for any
  $\grPprime$ and $\grQprime$ such that
  $\grPprime \leq \grQprime\plr{\rho.\gr\Psi}$, we have a function
  $\mathrm{lift}_\rho :
  \forallb{\V\,\grQprime\delta \dotto \W\,\grPprime\gamma}$,
  we can map environments of type $\Delta \env\V \Theta$ into environments of
  type $\Gamma \env\W \Theta$.
\end{lemma}
\begin{proof}
  Let $\rho$ be as in the statement, and let $\sigma : \Delta \env\V \Theta$.
  For the environment we are constructing, let
  $\gr\Psi \coloneqq \sigma.\gr\Psi; \rho.\gr\Psi$, noting that
  $\grP \leq \grQ\plr{\rho.\gr\Psi} \leq
  \plr{\grR\plr{\sigma.\gr\Psi}}\plr{\rho.\gr\Psi}$.
  For the action on variables, we are given $\grPprime \leq \grRprime\gr\Psi$
  with $\grRprime\theta \sqni A$.
  We can immediately apply the action of $\sigma$, giving us a value of type
  $\V\,\plr{\grRprime\plr{\sigma.\gr\Psi}}\,A$.
  We note that
  $\grPprime \leq \plr{\grRprime\plr{\sigma.\gr\Psi}}\plr{\rho.\gr\Psi}$, and
  apply $\mathrm{lift}_\rho$ to get the desired value.
\end{proof}

\begin{lemma}[Composition of environments]\label{thm:env-comp}
  Given a function
  $\mathrm{lift} : \plr{\rho : \grP\gamma \env\U \grQ\delta} \to
  \forall \grPprime, \grQprime.~\grPprime \leq \grQprime\plr{\rho.\gr\Psi} \to
  \forallb{\V\,\grQprime\delta \dotto \W\,\grPprime\gamma}$, then we can
  compose environments of types $\Gamma \env\U \Delta$ and
  $\Delta \env\V \Theta$ into an environment of type $\Gamma \env\W \Theta$.
\end{lemma}

\begin{corollary}\label{thm:env-postren}
  Given an environment $\rho : \Gamma \env\U \Delta$ and a renaming
  $\sigma : \Delta \env\sqni \Theta$, we can get an environment of type
  $\Gamma \env\U \Theta$.
  When $\U = {\sqni}$, this gives composition of renamings.
\end{corollary}
\begin{proof}
  By \cref{thm:env-comp}, with $\mathrm{lift}$ being given exactly by the
  action of $\rho$ on variables.
\end{proof}

\todo{What do these choices mean?}
\begin{example}
  We can derive the following instances of environment composition.
  \begin{itemize}
    \item If $\V = \W = {\vdash}$, then $\mathrm{lift}$ is given by a
      syntactic traversal.
      For example, if $\U = {\sqni}$, we need the action of renaming on terms
      to show that a renaming followed by a substitution composes to a
      substitution.
      If $\U = {\vdash}$, then the action of substitution on terms gives us that
      substitutions compose.
    \item More generally, if $\V = {\vdash}$ and we have a semantics from
      $\U$ to $\W$, then $\mathrm{lift}$ can be given by the semantic traversal
      of terms.
  \end{itemize}
\end{example}
